---
title: "VIM package: imputation examples"
output: html_document
date: "2025-06-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
```

### Install and load library

The `transformerImpute()` function sits inside the transformerImpute branch in the VIM github repository (<https://github.com/statistikat/VIM/tree/transformerImpute>). We install the VIM version respective to the transformerImpute branch and load it.

```{r loadVIM}
# load VIM package (transformer branch)
devtools::install_github("statistikat/VIM", ref = "transformerImpute")
library(VIM)
```

### Test data

To demonstrate our imputation methods, we use the Iris data set.

```{r iris}
dim(iris)
str(iris)
summary(iris)
```

#### Create missing values

We artificially inject missing values into the data, following three different missing mechanisms:

1.  Missing Completely At Random **(MCAR)**: The probability of a value missing is completely unrelated to any data.

2.  Missing At Random **(MAR)**: The probability of missingness depends on observed variables, but not the missing variable itself.

3.  Missing Not At Random **(MNAR)**: The probability of missingness depends on the missing value itself.

For this, we implemented a function for each missing mechanism (see `create_missings_function.R`)

The output of these functions will be the input data set, with missing value in the `miss_var`. Addditinoally we add the column `<miss_var>_TRUE`, which saves the original column values before missing value injection.

In addition to the parameters `dat` (input data), `miss_var` (variable to inject missings to) and `rate_missing` (rate of values missing), the functions `mar()` and `mnar()` take as argunemt `tab_list`, which is a named list of data tables giving the probability of non-response according to each variable value in the variable. The default non-response value is always 1. For variable values that are more likely to cause missingness, the non-response value will be \>1.

##### MCAR

First we start with MCAR. The missing variable is *Species.*

```{r}
source("create_missings_functions.R")
iris <- as.data.table(iris)
iris_mcar <- mcar(iris, "Species",rate_missing = 0.2)
sapply(iris_mcar,function(x) sum(is.na(x)))
```

##### MAR

Next we inject MAR. For this, we first have to create the `tab_list`. For this, we say that missing values for Species is more likely for larger *Sepal.Width* and lower values of *Petal.Length*.

```{r}
#find distinct values and counts
tab_Petal.Length <- iris[,.N,by=Petal.Length]
# default non_res value is 1
tab_Petal.Length[,non_res:=1]
# for values < 3, the non response increases
tab_Petal.Length[Petal.Length<3, non_res:=1.5]
tab_Petal.Length[Petal.Length<1.5,non_res:=2]

tab_Sepal.Width <- iris[,.N,by=Sepal.Width]
tab_Sepal.Width[,non_res:=1]
tab_Sepal.Width[Sepal.Width>3.5,non_res:=1.3]
tab_Sepal.Width[Sepal.Width>=4,non_res:=1.6]

tab_list <- list(Petal.Length = tab_Petal.Length,
                 Sepal.Width = tab_Sepal.Width)
tab_list
```

```{r}
iris_mar <- mar(iris, "Species", tab_list, rate_missing = 0.2)
```

##### MNAR

Lastly, we look at MNAR. We reuse the `tab_list` from the MAR generation, but add higher non-response for *Species* = setosa.

```{r}
tab_Species <- iris[,.N,by=Species]
tab_Species[,non_res:=1]
tab_Species[Species=="setosa",non_res:=1.5]
tab_list$Species <- tab_Species
tab_list$Species
```

```{r}
iris_mnar <- mnar(iris, "Species", tab_list, rate_missing = 0.2)
```

### Imputation

#### Transformer

We use the `transformerImpute()` function from the VIM transformerImpute branch to impute the missing values for all three data sets. For this, we leave all model hyperparameter as their default values.

Note that at this point, the function can only impute one variable at a time.

```{r impmcar, warning=FALSE}
imp_mcar <- transformerImpute(data = iris_mcar,
                              target = "Species", #variable to impute
                              cat_vars = "Species", #categorical variables
                              imp_col = TRUE, #add an imputation indicator column
                              include_cols = c("Sepal.Length","Sepal.Width","Petal.Length", "Petal.Width","Species"),
                              verbose = 0, #turn of training updates
                              epochs = 50)

# add the true species values
imp_mcar[,Species_TRUE:=iris_mcar$Species_TRUE]
```

```{r impmar, warning=FALSE}
imp_mar <- transformerImpute(data = iris_mar,
                              target = "Species", #variable to impute
                              cat_vars = "Species", #categorical variables
                              imp_col = TRUE, #add an imputation indicator column
                              include_cols = c("Sepal.Length","Sepal.Width","Petal.Length", "Petal.Width","Species"),
                             verbose = 0,
                             epochs = 50)

# add the true species values
imp_mar[,Species_TRUE:=iris_mar$Species_TRUE]
```

```{r impmnar, warning=FALSE}
imp_mnar <- transformerImpute(data = iris_mnar,
                              target = "Species", #variable to impute
                              cat_vars = "Species", #categorical variables
                              imp_col = TRUE, #add an imputation indicator column
                              include_cols = c("Sepal.Length","Sepal.Width","Petal.Length", "Petal.Width","Species"),
                              verbose = 0,
                              epochs = 50)

# add the true species values
imp_mnar[,Species_TRUE:=iris_mnar$Species_TRUE]
```

### Evaluation
Now we evaluate the performance and compare them for the different missing mechanism 
```{r evaluate}
dat_eval <- data.table(orig_values = iris$Species, 
                       value_mcar = imp_mcar$Species, imp_mcar = imp_mcar$imp,
                       value_mar = imp_mar$Species, imp_mar = imp_mar$imp,
                       value_mnar = imp_mnar$Species, imp_mnar = imp_mnar$imp)

dat_eval[,acc_mcar:=0]
dat_eval[,acc_mar:=0]
dat_eval[,acc_mnar:=0]

dat_eval[imp_mcar==1&orig_values==value_mcar,acc_mcar:=1]
dat_eval[imp_mar==1&orig_values==value_mar,acc_mar:=1]
dat_eval[imp_mnar==1&orig_values==value_mnar,acc_mnar:=1]

dat_eval[imp_mcar==0, acc_mcar:=NA]
dat_eval[imp_mar==0, acc_mar:=NA]
dat_eval[imp_mnar==0, acc_mnar:=NA]

mean(dat_eval$acc_mcar,na.rm = T)
mean(dat_eval$acc_mar,na.rm = T)
mean(dat_eval$acc_mnar,na.rm = T)
```






